\setcounter{section}{5} % Set to appropriate section number for Week 6

\section{Convexity}

\begin{definition*}[Convex Sets and Functions]
\begin{itemize}
    \item A subset $A \subseteq \R^n$ is \textbf{convex} if for any $x, y \in A$ and any $t \in [0,1]$, the point $p = (1-t)x + ty$ is also an element of $A$. \textcolor{eGray}{(The straight line segment connecting any two points in the set lies entirely within the set.)}
    
    \item A function $f: A \to \R$ is \textbf{convex} if its domain $A$ is a convex set, and for all $x, y \in A$ and $t \in [0,1]$:
    \[ f((1-t)x + ty) \le (1-t)f(x) + t f(y) \]
\end{itemize}
\end{definition*}

\gemininote{The functional inequality simply states that if you draw a straight secant line between two points on the graph of $f$, the actual graph of the function will always hang \emph{below} (or exactly on) that secant line. It looks like a bowl.}

\begin{proposition*}
Let $U \subseteq \R^n$ be open and convex, and $f \in C^2(U)$. The following statements are perfectly equivalent:
\begin{enumerate}
    \item $f$ is a convex function.
    \item The Hessian matrix $\mathscr{H}f(x)$ is positive semi-definite (non-negative definite) for all $x \in U$.
    \item $f(y) - f(x) \ge Df_x(y-x)$ for all $x, y \in U$. \textcolor{eGray}{(The graph of the function always lies above any of its tangent planes.)}
\end{enumerate}
\end{proposition*}

\begin{example*}
The $n$-dimensional parabola $p: \R^n \to \R, x \mapsto \abs{x}^2 = \sum_{i=1}^n x_i^2$ is convex. 
Taking the second derivatives, we get $\partial_i \partial_j p(x) = 2\delta_{ij}$ (which is $2$ if $i=j$, and $0$ otherwise). Thus, the Hessian is strictly diagonal:
\[ \mathscr{H}p(x) = 2 I_n \]
Every eigenvalue of this matrix is $2 > 0$, so the matrix is strictly positive definite. This immediately implies the function is convex!
\end{example*}

\hrulefill

\begin{exercise*}
Let $U \subseteq \R^n$ be open and convex, and $f \in C^1(U)$.
\begin{enumerate}
    \item Show that $f: U \to \R$ is convex if and only if the 1-dimensional slice functions $f_{x,y}: [0,1] \to \R, s \mapsto f(x + s(y-x))$ are convex for all $x, y \in U$.
    \item Use this to prove the tangent plane property: $f$ convex $\iff f(y) - f(x) \ge Df_x(y-x)$.
    \item Conclude that if a convex function $f$ has a critical point at $x_0 \in U$, then it has a \emph{global} minimum at $x_0$.
\end{enumerate}
\end{exercise*}

\begin{solution}\hfill\\
\textbf{a) The 1D Reduction} \\
Let $f_{x,y}$ be convex. We want to show $f$ is convex.
\begin{equation}
\begin{aligned}
    f_{x,y}((1-t)s_0 + t s_1) &\le (1-t)f_{x,y}(s_0) + t f_{x,y}(s_1) \\
    f(x + [(1-t)s_0 + t s_1](y-x)) &\le (1-t)f(x + s_0(y-x)) + t f(x + s_1(y-x)) 
\end{aligned}
\end{equation}

If we simply set $s_0 = 0$ and $s_1 = 1$, the right side simplifies to $(1-t)f(x) + t f(y)$, and the left side simplifies exactly to $f((1-t)x + ty)$. This immediately proves $f$ is convex!
Vice versa, if $f$ is convex, the convexity of $f_{x,y}$ follows by applying the convexity of $f$ to the two points $u = x + s_0(y-x)$ and $v = x + s_1(y-x)$.

\vspace{0.3cm}
\textbf{b) Tangent Plane Inequality} \\
Recall from Analysis 1 that a 1D function $g \in C^1(I)$ is convex if and only if its derivative $g'$ is monotonically increasing. 
By part (a), $f$ is convex $\iff f_{x,y}$ is convex $\iff f_{x,y}'$ is increasing $\forall x,y \in U$.

Compute the derivative of the slice using the chain rule for $t \in [0,1]$:
\[ f_{x,y}'(t) = Df_{x + t(y-x)}(y-x) \]
By the 1D Mean Value Theorem, there must exist some $s \in (0,1)$ such that:
\[ f(y) - f(x) = f_{x,y}(1) - f_{x,y}(0) = f_{x,y}'(s) = Df_{x + s(y-x)}(y-x) \]
Now we apply the increasing property. If $f$ is convex, then $f_{x,y}'(s) \ge f_{x,y}'(0)$.
\[ f(y) - f(x) = f_{x,y}'(s) \ge f_{x,y}'(0) = Df_x(y-x) \]
\textcolor{eGray}{(The proof for the reverse direction involves evaluating the inequality for intermediate points $u$ and $v$ to show that $f_{x,y}'$ must be increasing. See the handwritten notes for the algebraic manipulation.)}

\vspace{0.3cm}
\textbf{c) Critical Points of Convex Functions} \\
Since $f$ is convex, we can always apply the inequality we just proved. If $x_0$ is a critical point, its differential is zero ($Df_{x_0} = 0$). Plugging this into the inequality for any arbitrary point $y \in U$:
\[ f(y) - f(x_0) \ge Df_{x_0}(y - x_0) = 0 \implies f(y) \ge f(x_0) \]
Since this holds for \emph{all} $y \in U$, $x_0$ is an absolute global minimum!
\end{solution}

\hrulefill

\section{The Inverse Function Theorem (IFT)}

\begin{theorem*}[Inverse Function Theorem]
Let $U \subseteq \R^n$ be open and $f \in C^k(U, \R^n)$. Let $x_0 \in U$ be a point such that the differential $Df_{x_0}$ is strictly invertible (i.e., $\det(Jf(x_0)) \ne 0$).

Then there exists an open neighborhood $U_0 \subseteq U$ around $x_0$ such that:
\begin{enumerate}
    \item $f|_{U_0}: U_0 \to f(U_0)$ is a $C^k$-diffeomorphism. \textcolor{eGray}{(It is locally invertible and smooth!)}
    \item The differential of the inverse is the inverse of the differential: 
    \[ D(f|_{U_0}^{-1})_{f(x)} = \left( D(f|_{U_0})_x \right)^{-1} \quad \forall x \in U_0 \]
\end{enumerate}
\end{theorem*}

\gemininote{Intuition: If the derivative matrix is invertible, it means the function locally acts like a linear transformation that scales or rotates space, but it never \qt{flattens} a dimension down to zero. Because no information is destroyed (no dimensions are crushed), you can easily reverse the process—at least locally!}

\begin{question*}
\begin{enumerate}
    \item If $f: \R^n \to \R^n$ is $C^k$ and a homeomorphism (bijective with continuous inverse), is it necessarily a $C^k$-diffeomorphism?
    \item If $f \in C^\infty(U, \R^n)$ has an everywhere invertible differential ($Df_x$ invertible $\forall x \in U$), is $f$ a $C^\infty$-diffeomorphism onto its image?
\end{enumerate}
\end{question*}

\begin{answer*}
\textbf{1. False.} Consider $f: \R \to \R, x \mapsto x^3$. This function is bijective and continuous, and its inverse $f^{-1}(x) = x^{1/3}$ is also continuous. However, $f^{-1}$ is not differentiable at $x=0$, so it is not a diffeomorphism!

\textbf{2. False.} Local invertibility does not guarantee global invertibility! Consider $f(x) = \cos(x)$ restricted to $\R \setminus \pi\mathbb{Z}$. For any $x$ in this domain, $f'(x) = -\sin(x) \ne 0$, so the differential is everywhere invertible. The IFT guarantees it is locally a diffeomorphism everywhere. But globally, cosine oscillates and hits the same values infinitely many times, so it is obviously not bijective onto $(-1, 1)$!
\end{answer*}

\hrulefill

\section{The Implicit Function Theorem}

\begin{theorem*}[Implicit Function Theorem]
\textbf{Setup:} Let $n > n-d \ge 1$. Let $f \in C^k(\R^d \times \R^{n-d}, \R^{n-d})$. Suppose $f(x_0, y_0) = 0$ for some point. 
Suppose $D_y f(x_0, y_0)$, the differential with respect to the $y$-variables, is bijective (i.e., $J_y f(x_0, y_0)$ is an invertible square matrix).

\textbf{Consequence:} There exist open balls $B_r(x_0) \subseteq \R^d$ and $B_s(y_0) \subseteq \R^{n-d}$, and a $C^k$ function $g: B_r(x_0) \to B_s(y_0)$ such that for all points in this local window:
\[ f(x,y) = 0 \iff y = g(x) \]
Furthermore, the Jacobian of this new implicit function is:
\[ Jg(x) = -\left( J_y f(x, g(x)) \right)^{-1} J_x f(x, g(x)) \]
\end{theorem*}

\gemininote{The TA wrote: \qt{The geometrical meaning of this formula eludes me, sorry.} But it's actually incredibly logical! It's just the Multivariable Chain Rule. We know that locally, $f(x, g(x)) = 0$. If we take the total derivative of both sides with respect to $x$, the Chain Rule gives:
$J_x f + J_y f \cdot Jg = 0$. 
Now just solve for $Jg$! Move $J_x f$ to the other side, and multiply from the left by the inverse of $J_y f$. That's it!}

\begin{example*}[1. The Unit Circle]
Consider $F(x,y) = x^2 + y^2 - 1 = 0$. Can we express $y$ as a function of $x$?
Taking the $y$-derivative: $D_y F(x,y) = 2y$.
This is invertible (non-zero) everywhere \emph{except} at $y=0$ (the points $(-1,0)$ and $(1,0)$). 
Thus, for any point where $y \ne 0$, the IFT guarantees we can locally write $y(x) = \pm\sqrt{1-x^2}$. At $y=0$, the tangent is perfectly vertical, so it fails the vertical line test and cannot be written as $y(x)$.
\end{example*}

\begin{example*}[2. The Quintic Equation]
Consider the level set $G^{-1}\set{0} = \set{(x,y) \in \R^2 : y^5 + y^3 + y + x = 0}$.
Clearly, the origin $(0,0)$ is a solution since $0^5 + 0^3 + 0 + 0 = 0$.
Check the partial derivatives at $(0,0)$:
\begin{equation}
\begin{aligned}
    \partial_x G(0,0) &= 1 \ne 0 \\
    \partial_y G(0,0) &= \left[ 5y^4 + 3y^2 + 1 \right]_{y=0} = 1 \ne 0
\end{aligned}
\end{equation}
Since $\partial_y G \ne 0$, the IFT guarantees that there exists an open interval around $x=0$ where we can uniquely write $y = g(x)$. 

\textcolor{eSaddleBrown}{Why is this theorem so powerful?} By the famous Abel-Ruffini Theorem from Galois theory, there is mathematically \textbf{no general algebraic formula} to solve a polynomial of degree 5. You literally cannot write down $y(x)$ using standard roots. Yet, the Implicit Function Theorem mathematically guarantees that this function $y(x)$ exists, that it is perfectly smooth, and we can even compute its derivative without ever knowing the function's explicit formula!
\end{example*}

\setcounter{section}{5} % Week 6

\section{Convexity}

\begin{definition*}[Convex Sets and Functions]
\begin{itemize}
    \item A subset $A \subseteq \R^n$ is \textbf{convex} if for any $x, y \in A$ and any $t \in [0,1]$, the point $p = (1-t)x + ty$ is also an element of $A$. \textcolor{eGray}{(Intuitively: The straight line segment connecting any two points in the set lies entirely within the set.)}
    
    \item A function $f: A \to \R$ is \textbf{convex} if its domain $A$ is a convex set, and for all $x, y \in A$ and $t \in [0,1]$:
    \[ f((1-t)x + ty) \le (1-t)f(x) + t f(y) \]
\end{itemize}
\end{definition*}

\begin{center}
\begin{tikzpicture}[scale=1.5, very thick]
    % Axes
    \draw[->, eBlack] (-0.5, 0) -- (3.5, 0) node[right] {$x$};
    \draw[->, eBlack] (0, -0.5) -- (0, 2.5) node[above] {$f(x)$};
    
    % Convex curve
    \draw[eMediumBlue, ultra thick] (0.5, 2) .. controls (1.5, 0) and (2, 0) .. (3, 1.5);
    
    % Points and Secant
    \coordinate (X) at (0.8, 1.25);
    \coordinate (Y) at (2.8, 1.25);
    \draw[eDarkRed, dashed] (X) -- (Y);
    \fill[eDarkRed] (X) circle (2pt) node[left] {$f(x)$};
    \fill[eDarkRed] (Y) circle (2pt) node[right] {$f(y)$};
    
    \node[eDarkRed, align=center] at (1.8, 1.6) {\small Secant line lies \\ \small \textbf{above} the graph};
\end{tikzpicture}
\end{center}

\gemininote{The functional inequality simply states that a secant line between any two points on the graph always stays \qt{above} the function itself. Think of a bowl-shaped graph.}

\begin{proposition*}
Let $U \subseteq \R^n$ be open and convex, and $f \in C^2(U)$. The following statements are equivalent:
\begin{enumerate}
    \item $f$ is a convex function.
    \item The Hessian matrix $\mathscr{H}f(x)$ is positive semi-definite for all $x \in U$.
    \item $f(y) - f(x) \ge Df_x(y-x)$ for all $x, y \in U$. \textcolor{eGray}{(The graph always lies above its tangent planes.)}
\end{enumerate}
\end{proposition*}

\hrulefill

\section{The Inverse Function Theorem (IFT)}

\begin{theorem*}[Inverse Function Theorem]
Let $U \subseteq \R^n$ be open and $f \in C^k(U, \R^n)$. Let $x_0 \in U$ be a point such that the differential $Df_{x_0}$ is invertible ($\det(Jf(x_0)) \ne 0$).

Then there exists an open neighborhood $U_0 \subseteq U$ around $x_0$ such that:
\begin{enumerate}
    \item $f|_{U_0}: U_0 \to f(U_0)$ is a $C^k$-diffeomorphism.
    \item The derivative of the inverse is the inverse of the derivative: 
    \[ D(f^{-1})_{f(x)} = (Df_x)^{-1} \]
\end{enumerate}
\end{theorem*}

\begin{question*}
\begin{enumerate}
    \item If $f: \R^n \to \R^n$ is a $C^k$ homeomorphism, is it necessarily a $C^k$-diffeomorphism?
    \item Does local invertibility everywhere imply global invertibility?
\end{enumerate}
\end{question*}

\begin{answer*}
\textbf{1. No.} Example: $f(x) = x^3$. The inverse $x^{1/3}$ is not differentiable at 0. \\
\textbf{2. No.} Example: $f(x) = (\cos x, \sin x)$ has invertible derivative everywhere but is periodic (not injective).
\end{answer*}

\hrulefill

\section{The Implicit Function Theorem}

\begin{theorem*}[Implicit Function Theorem]
Suppose $f(x_0, y_0) = 0$ and the partial Jacobian $J_y f(x_0, y_0)$ is invertible. Then there exists a unique local function $y = g(x)$ such that $f(x, g(x)) = 0$.
The derivative is given by:
\[ Jg(x) = -(J_y f)^{-1} J_x f \]
\end{theorem*}

\gemininote{This formula is just the Chain Rule in disguise! Differentiating $f(x, g(x)) = 0$ with respect to $x$ gives $J_x f + J_y f \cdot Jg = 0$. Solving for $Jg$ yields the result.}

\begin{example*}[The Unit Circle]
Consider $f(x,y) = x^2 + y^2 - 1 = 0$. We check $\partial_y f = 2y$.
This is invertible if $y \ne 0$. Thus, for any point not on the x-axis, we can locally solve for $y = \pm\sqrt{1-x^2}$.

\begin{center}
\begin{tikzpicture}[scale=1.5, very thick]
    \draw[eMediumBlue] (0,0) circle (1cm);
    \draw[->, eBlack] (-1.5,0) -- (1.5,0) node[right] {$x$};
    \draw[->, eBlack] (0,-1.2) -- (0,1.2) node[above] {$y$};
    
    % Points where it fails
    \fill[eDarkRed] (1,0) circle (1.5pt);
    \fill[eDarkRed] (-1,0) circle (1.5pt);
    \draw[eDarkRed, <->] (1.2, -0.3) -- (1.2, 0.3) node[right] {$\partial_y f = 0$};
\end{tikzpicture}
\end{center}
\geminiwarning{At $y=0$, the tangent is vertical. The theorem fails here because a small step in $x$ would not uniquely determine $y$ (it could go up or down).}
\end{example*}

\section{The Implicit Function Theorem}

\begin{theorem*}[Implicit Function Theorem]
\textbf{Setup:} Let $n > n-d \ge 1$. Let $f \in C^k(\R^d \times \R^{n-d}, \R^{n-d})$. Suppose $f(x_0, y_0) = 0$ for some point $(x_0, y_0) \in \R^d \times \R^{n-d}$[cite: 16]. 
Suppose the partial differential with respect to the $y$-variables, $D_y f(x_0, y_0)$, is bijective (i.e., the Jacobian $J_y f(x_0, y_0)$ is an invertible square matrix)[cite: 16].

\textbf{Consequence:} There exist open neighborhoods $U \subseteq \R^d$ of $x_0$ and $V \subseteq \R^{n-d}$ of $y_0$, and a unique $C^k$ function $g: U \to V$ such that for all $(x,y) \in U \times V$:
\[ f(x,y) = 0 \iff y = g(x) \]
Furthermore, the derivative of this implicit function is given by:
\[ Jg(x) = -\left( J_y f(x, g(x)) \right)^{-1} J_x f(x, g(x)) \]
\end{theorem*}

\gemininote{The TA notes mentioned that the geometrical meaning of this derivative formula \qt{eludes} them[cite: 16]. But it is actually a beautiful application of the \textbf{Multivariable Chain Rule}! Since we know $f(x, g(x)) = 0$ for all $x \in U$, differentiating both sides with respect to $x$ gives:
\[ J_x f + J_y f \cdot Jg = 0 \]
Solving for $Jg$ by moving $J_x f$ and multiplying by the inverse of $J_y f$ leads directly to the formula!}

\begin{example*}[1. The Unit Circle $S^1$]
Consider the level set of $f(x,y) = x^2 + y^2 - 1 = 0$[cite: 16]. We want to know if we can locally solve for $y$ as a function of $x$.
The partial derivative with respect to $y$ is $\partial_y f = 2y$[cite: 16]. This is invertible (non-zero) as long as $y \neq 0$[cite: 16].

\begin{center}
\begin{tikzpicture}[scale=2, very thick]
    % Unit circle
    \draw[eMediumBlue] (0,0) circle (1cm);
    \draw[->, eBlack] (-1.5, 0) -- (1.5, 0) node[right] {$x$};
    \draw[->, eBlack] (0, -1.2) -- (0, 1.2) node[above] {$y$};
    
    % Point (x0, y0)
    \coordinate (P) at (0.6, 0.8);
    \fill[eDarkGreen] (P) circle (1.5pt) node[above right] {$(x_0, y_0)$};
    
    % Local window U x V
    \draw[eDarkGreen, dashed] (0.3, 0.6) rectangle (0.9, 1.0);
    \node[eDarkGreen] at (0.3, 1.1) {\small Window $U \times V$};
    
    % Highlighted local function y = g(x)
    \draw[eDarkRed, ultra thick] (0.3, 0.954) arc (107.5:64.2:1cm);
    \node[eDarkRed] at (1.1, 0.9) {\small $y = \sqrt{1-x^2}$};

    % Points where it fails (y=0)
    \fill[eDarkRed] (1,0) circle (1.5pt);
    \fill[eDarkRed] (-1,0) circle (1.5pt);
    \draw[eDarkRed, <->] (1.15, -0.2) -- (1.15, 0.2) node[midway, right, align=left] {\small Vertical tangent \\ \small ($\partial_y f = 0$)};
    \node[eDarkRed] at (0, -0.4) {\small \qt{not possible around $y=0$}};
\end{tikzpicture}
\end{center}

\geminiwarning{Around the points $(1,0)$ and $(-1,0)$, the tangent to the circle is vertical[cite: 16]. In any small window around these points, a single $x$-value would correspond to two different $y$-values (one on the upper half, one on the lower half). Therefore, we cannot define a unique function $y = g(x)$ there, which matches the fact that $\partial_y f = 0$ at those points.}
\end{example*}

\begin{example*}[2. The Quintic Equation]
Consider $G(x,y) = y^5 + y^3 + y + x$[cite: 16]. Does the level set $G^{-1}\{0\}$ define a function $y = g(x)$ around $(0,0)$?
We check the solution: $G(0,0) = 0^5 + 0^3 + 0 + 0 = 0$.
Now check the $y$-derivative: $\partial_y G = 5y^4 + 3y^2 + 1$[cite: 16].
At $(0,0)$, we have $\partial_y G(0,0) = 1 \neq 0$[cite: 16].

By the Implicit Function Theorem, a local function $y = g(x)$ exists! This is a massive result because, according to Abel's Theorem, there is no general formula to solve for the roots of a 5th-degree polynomial. Yet, we know $y(x)$ exists and is smooth.
\end{example*}